Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-07 paper/paper.tex
a pass on rl sections

\subsubsection{Efficient learning via value approximation and experience replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to efficiently approximate the state-action value function
to allow for scaling to large state and action spaces.
In a state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
At the end of this cycle, a new \textit{experience} $e = (s,a,r,s')$ is produced, which represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
$y := r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$. Here the function $(s,a) \mapsto Q(a,s|\theta^Q_{k})$
% \textbf{XXX just double-checking, are the dots int he Q() correct?}
is an approximator for the $Q$-value operator,  parameterized by weights $\theta^Q_{k+1}$
(synaptic strengths, etc.).

@@

Computing an optimal policy then corresponds to finding the parameters $\theta^Q_{k+1}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k+1})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q_{k+1}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal U(D)}\left[\frac{1}{2}(Q(s, a|\theta^Q_{k+1}) - y)^2\right],
  \label{eq:oracle}
\end{equation}
where
$ y: = r + \gamma \max_{a' \in \mathcal A} Q(s', a'|\theta^Q_k) = r + \gamma Q(s', \pi(s')|\theta^Q_k)$ is the prediction target.
% While $y_t$ is also dependent on $\theta^Q$, this is typically ignored.
For instance, a general linear model with a kernel $\phi$ would be of the
form
$${Q}(s, a|\theta^Q) = \phi(s,a)^T\theta^Q.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016}.
%  and Parr et al. 2003 (JMLR)].
A recently proposed, practically successful alternative approach
\citep{mnih2015,silver2016mastering} is to learn this
representation using a deep neural network, leading to the so-called \textit{Deep Q-learning} family of methods which
likely are state-of-the-art in RL.
The set of $\theta^Q$ parameters that instantiate the nonlinear interactions
between layers of the artifical neural network
may find a neurobiological correspondence in the adaptive strengths of axonal
connections between neurons from the different levels
of the neural processing hierarchy
\citep{mesulam1998sensation, taylor2015global}.

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta \theta_{k+1}^Q \propto -\nabla_{\theta^Q_{k+1}}\mathcal L(\theta^Q_{k+1})
  % = -\mathbb E_{(s, a, r, s') \sim U(\mathcal D)}[(Q(s, a|\theta^Q) - Q(s',a'|\theta^Q_{}))^2]
  = -\mathbb E_{(s, a, r, s') \sim \mathcal U(D)}\left[\underbrace{(Q(s, a|\theta^Q_{k+1}) - y)}_{\text{regret / RPE}}
    \underbrace{\nabla_{\theta^Q_{k}}Q(s, a|\theta^Q_{k+1})}_{\text{averseness}}\right].
  \label{eq:oracle}
\end{equation}

\subsubsection{Link to classical reinforcement learning algorithms} One should note that most classical RL algorithms, including \textit{Temporal Difference (TD)}
learning \citep{sutton1998reinforcement}, REINFORCE \citep{williams1992}, and SARSA can be
cast in this \textbf{please unpack the "this" here}
general variational framework. For instance, TD corresponds to
the above framework using a linear value approximator with feature encoding
$\phi(s,a) = \delta_{(s,a)} =  \text{ point mass at }(s,a)$ on the grid
$\mathcal S \times \mathcal A$, and so
$$\nabla_{\theta^Q_{k+1}}Q(s, a|\theta^Q_{k+1}) = \phi(s, a) = \delta_{(s,a)}.$$
Whence, the gradient update due to the sample $(s,a,r,s') \in \mathcal D$ is
$$\theta^Q(s,a) \leftarrow \theta^Q(s,a) + (Q(s, a|\theta^Q_{k+1}) - y),$$
the well-known TD update rule \citep{sutton1998reinforcement}.

@@

While sensory experience is a generative process from both views,
the predictive coding account frames
sensory perception of the external world
as a generative experience due to the modulatory top-down influence at
various hierarchical levels of sensory input processing. Instead,
the Markov property of our sequential decision making framework.
In our RL view, the environmental model is
\textbf{effectively expressed in the current state of the human agent
XXX guys, please double-check}.
The hierarchical processing aspect from predictive coding
is re-expressed in our account in form of
nested prediction of probable upcoming actions, states, and rewards.
Both frameworks model the consequences of action.
The predictive coding framework is typically explained without
explicitly parameterization of the agent's horizon and
has a tendency to be presented as emphasizing prediction about the
immediate future.
In the present account, the horizon of that
look into the future is explicitly expressed in the $\gamma$ parameter
of the Bellman equation.
% \item \suggestremove{Mismatch negativity of PC is immanent in reward prediction error
% by the difference between actually predicted reward of an action given
% a state and the reward predicted by non-linear regression with graded
% discounting of future rewards.}
The process of adapting the neuronal connections
for improved top-down modulation
takes the concrete form of gradient computation and
back-propagation in our RL implementation.
It is however important to note that
the neurobiological plausibility of
the back-propagation procedure is controversal
\citep{goodfellow2016deep}.
\item In sum,
the formal account may serve as a concrete implementation of
the predictive coding account
from cognitive neuroscience.
MPDs have the advantage of exposing an explicit
mechanism of the horizon of future considerations or
how the internal model of the world gets updated,
and
more explicitly reconciles sensory input processing and action output preparation.
%
As such, recasting DMN function in terms of MDPs naturally incorporates
a majority of aspects of the prediction coding hypothesis.

>>>>

\subsubsection{Efficient learning via value approximation and experience replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to efficiently approximate the state-action value function
to allow for scaling to large state and action spaces.
In a state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
At the end of this cycle, a new \textit{experience} $e = (s,a,r,s')$ is produced, which represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := r_k + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the output of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$. Here the function $(s,a) \mapsto Q(a,s|\theta_{k-1})$
% \textbf{XXX just double-checking, are the dots int he Q() correct?}
is an approximator for the $Q$-value operator,  parameterized by weights $\theta_{k}$
(synaptic strengths, etc.).

@@

Computing an optimal policy then corresponds to finding the parameters $\theta_{k}$ which minimize the following mean-squared loss function
\begin{equation}
  \mathcal L(\theta^Q_{k})
  % = \mathbb E_{(s, a, r, s') \sim \mathcal D}[(\tilde{Q}(s, a|\theta) - \tilde{Q}(s',a'|\theta^Q_{k}))^2]
  = \mathbb E_{(s, a, r, s') \sim \mathcal D}\left[\frac{1}{2}(\tilde{Q}(s, a|\theta_{k}) - y_k)^2\right].
  \label{eq:oracle}
\end{equation}
For instance, a general linear model with a kernel $\phi$ would be of the
form
$$\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta.$$
$\phi(s,a)$ would represent a high-level hand-crafted representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016}.
%  and Parr et al. 2003 (JMLR)].
A recently proposed, practically successful alternative approach
\citep{mnih2015,silver2016mastering} is to learn this
representation using a deep neural network, leading to the so-called \textit{Deep Q-learning} family of methods which
likely are state-of-the-art in RL.
The set of $\theta^Q$ parameters that instantiate the nonlinear interactions
between layers of the artifical neural network
may find a neurobiological correspondence in the adaptive strengths of axonal
connections between neurons from the different levels
of the neural processing hierarchy
\citep{mesulam1998sensation, taylor2015global}.

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta \theta_{k} \propto -\nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\mathbb E_{(s, a, r, s') \sim \mathcal U(D)}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{averseness}}].
  \label{eq:oracle}
\end{equation}

\subsubsection{Link to classical reinforcement learning algorithms} One should note that most classical RL algorithms, including \textit{Temporal Difference (TD)}
learning \citep{sutton1998reinforcement}, REINFORCE \citep{williams1992}, and SARSA can be
cast in this \textbf{please unpack the "this" here}
general variational framework. For instance, TD corresponds to
the above framework using a linear value approximator with feature encoding
$\phi(s,a) = \delta_{(s,a)} =  \text{ point mass at }(s,a)$ on the grid
$\mathcal S \times \mathcal A$, and so
$$\nabla_{\theta}\tilde{Q}(s, a|\theta) = \phi(s, a) = \delta_{(s,a)}.$$
Whence, the gradient update due to the sample $(s,a,r,s') \in \mathcal D$ is
$$\theta(s,a) \leftarrow \theta(s,a) + \alpha(\tilde{Q}(s, a|\theta) - y),$$
the well-known TD update rule \citep{sutton1998reinforcement}.

@@

While sensory experience is a generative process from both views,
the predictive coding account frames
sensory perception of the external world
as a generative experience due to the modulatory top-down influence at
various hierarchical levels of sensory input processing. Instead,
the Markov property of our sequential decision making framework.
In our RL view, the environmental model is
\textbf{effectively expressed in the current state of the human agent
XXX guys, please double-check}.
The hierarchical processing aspect from predictive coding
is re-expressed in our account in form of
nested prediction of probable upcoming actions, states, and rewards.
Both frameworks model the consequences of action.
The predictive coding framework is typically explained without
explicitly parameterization of the agent's horizon and
has a tendency to be presented as emphasizing prediction about the
immediate future.
In the present account, the horizon of that
look into the future is explicitly expressed in the $\gamma$ parameter
of the Bellman equation.
% \item \suggestremove{Mismatch negativity of PC is immanent in reward prediction error
% by the difference between actually predicted reward of an action given
% a state and the reward predicted by non-linear regression with graded
% discounting of future rewards.}
The process of adapting the neuronal connections
for improved top-down modulation
takes the concrete form of gradient computation and
back-propagation in our RL implementation.
It is however important to note that
the neurobiological plausibility of
the back-propagation procedure is controversal
\citep{goodfellow2016deep}.
In sum,
the formal account may serve as a concrete implementation of
the predictive coding account
from cognitive neuroscience.
MPDs have the advantage of exposing an explicit
mechanism of the horizon of future considerations or
how the internal model of the world gets updated,
and
more explicitly reconciles sensory input processing and action output preparation.
%
As such, recasting DMN function in terms of MDPs naturally incorporates
a majority of aspects of the prediction coding hypothesis.
